---
title: "Filtering Humdrum Data"
author: "Nathaniel Condit-Schultz"
date:   "July 2022"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Filtering Humdrum Data}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: sentence
---

```{r, include = FALSE, message=FALSE, echo = FALSE }
knitr::opts_chunk$set(
  collapse = TRUE, message = FALSE,
  comment = ">\t"
)
library(humdrumR)
```

Most humdrum datasets contain a wealth of information: rhythms, pitches, lyrics, etc.
However, the particular analysis you want to conduct today might not need all that information.
The first step after [loading humdrum data](ReadWrite.html "Reading and writing humdrum data")
is usually to "filter out" the information you don't need.
(Of course, you are not erasing it from the original files, which are safe on your computer!)
This is often just one part of larger task of [preparing your data](Reshaping.html "Shaping humdrum data").
You don't necessarily *need* to do this, but usually it makes things simpler when you simply get anything you don't
need out of the way.

In this article, we'll use our pre-packaged Bach chorale data to illustrate how filtering works,
so let's load it up:

```{r message=FALSE}
chorales <- readHumdrum(humdrumRroot, 'HumdrumData/BachChorales/.*krn')

```

# Indexing

The simplest, common, way to filter your data is through *indexing*, which is a common approach in
programming for selecting subsets of data.
In `R`, we use the square brackets, `[]`, to index various data objects,
including basic atomic vectors, data.tables, and lists.
For example, using brackets I can extract subsets of a vector:

```{r}

myname <- c('N', 'A', 'T', ' ', 'C', 'O', 'N', 'D', 'I', 'T')

myname[1:3]

myname[5:10]

myname[c(1, 5)]

```


Watch out! `R` actually has **two** different ways of indexing:
you can use a single pair of matches brackets (`[ ]`) or a double pair (`[[ ]]`).
If you want a primer on how these two types of indexing are used in normal `R`
objects, check out the [R primer](Rbasics.html "Basics of R")---you really don't *need*
to do this before getting into `humdrumR`, but you might want to eventually.

We can use the single-bracket `[]` or double-bracket `[[]]` commands to index our `humdrumR` data objects.
The single-brackets are used to index whole *pieces* in your data.
The double-brackets are used to index "within" pieces in your data.
Either type of bracket can accept either `numeric` or `character` vectors to index by.

## Single-bracket indexing


The single-brackets are used to index whole *pieces* in your data.
So if you have a dataset of 19 files, but you only want to look at some of those files, you'd use
 the single-brackets `[]`.
 
### Numeric indices (for single-brackets)

If we give a `numeric` value `i` to a single-bracket index command, that number will select the `i`th 
file in the dataset.
For example, if we want to look only at the fifth chorale, we can call:

```{r}

chorales[5]


```

We can also give the command a vector of `i` numbers:

```{r}

chorales[c(1,3,5)]

```

You might want to use the `R` sequence command, `:`, to select a range of numbers:

```{r}

chorales[6:10]

```

### Character indices (for single-brackets)

If you supply a `humdrumR` object single-bracket index a `character` string,
`humdrumR` will treat that string as a regular expression and return all the files that
contain a match to that expression in *any* data token---even if there is only one.

For example, we might be only interested in files that use flat notes.
Since `**kern` represents flats with "`-`", we can simply write:

```{r}


chorales['-']

```

Look at that, five of our ten chorales contain *at least one* flat.
The other chorales have zero flats.
Notice that we still get the *whole* files returned to us!

## Double-bracket indexing

The double-brackets are used to index with pieces in your data.
Specifically, if you only want to work with certain spines or records within your pieces.
The double-brackets apply the same filters separately to each file in the dataset.

In double-bracket indexing, you can provide two separate arguments, either individually or together:

+ `i` is the first argument which will index rows *within* each file.
+ `j` is the second argument, which will index spines *within* each file.

If you want to use the `j` (spine) argument by itself, you should put a comma before it, to indicate that you are
skipping over `i`.
It is also good practice to by clear and put a comma after `i`, but this is not actually needed.
Basically, commands should look like this:

+ `chorales[[i, ]]` (records)
+ `chorales[[  , j]]` (spines)
+ `chorales[[i , j]]` (records *and* spines)


### Numeric indices (for double-brackets)

`Numeric` values `i` *or* `j` can be given to double-bracket `humdrumR` index commands.
For `i`, the number is simply matched to record numbers in each file.
For example, you could extract the first fifty records from each file as so:

```{r}

chorales[[1:50, ]]

```


For `j`, the number is matched to the spines in each file (from left to right).
So if you wanted the second spine in each file, you'd write:

```{r}
chorales[[ , 2]]

```

If you give indices that are larger than the number of records or spines in a file,
that file will be dropped.
For example, if we call

```{r}

chorales[[150:200, ]]

```

We only get three files back, because the other seven files don't have any records above 150!
(Notice that `humdrumR` won't remove the Exclusive interpretation or spine spine closing (`*-`) records...since that would break the humdrum syntax.)


### Character indices (for double-brackets)

`Character` string values `i` *or* `j` can also be given to double-bracket `humdrumR` index commands.
`humdrumR` will treat string as a regular expression and return all records (`i`) or spines (`j`)
that any match to that expression in a data token---even if there is only one.
For example, let's (again) say we are interesting in studying flats.
Since `**kern` represents flats as "`-`", we can find records that contain a flat by calling:


```{r}

chorales[['-', ]]

```

As with our single-bracket search for flats (previous section) we only get five files back,
because only five of the chorales contain *any* flats at all.
However, we now see that all records that don't contain a flat are completely stripped away,
leaving a (handful) of records with at least one flat.

If we do the same thing with `j` (spines)

```{r}
chorales[[ , '-']]

```

It looks like we get only one spine in each of our five files.
Those single spines are simply which ever spine in *each* file contained *any* flat.
But a couple things to notice!
In the first file, what is now spine `1` was the old spine `2`, which we can tell because that 
instrument interpretation is `*Itenor`.
However, if you look at the last of the five files

```{r}

chorales[[ , '-']][5]

```

we see `*Ibass`---so the new spine `1` was actually the original spine `1` in that file.
And wait, there's more! If we dig in and look at the other files returned to us (which aren't shown by default),

```{r}
chorales[[ , '-']][3]

```

we see that what is now the third file actually had flats in all four spines, and so all four spines are still there!

---

In some cases you may find the renumbering of spines on a *per file* basis confusing, and you want to keep track of the original spine numbers.
Fortunately, for you there is an option that might helpful, which you will learn more about below: 
try specifiying `removeEmpty = FALSE`:

```{r}
chorales[[ , '-', removeEmpty = FALSE]]

```


## Negative numeric indices

A nifty feature of `R` is that if you supply negative numbers to an indexer, `R` will *remove* those
numbers.
This works in `humdrumR` too, so if you want all the files *except* the first file, you could write:

```{r}
chorales[-1]

```

Or if you want all the spines *except* the fourth spine, write

```{r}

chorales[[ , -4]]

```

Or if you want to remove the first 20 records from each file:


```{r}

chorales[[-1:-20, ]]

```

(Again, `humdrumR` won't remove the Exclusive interpretation or spine spine closing (`*-`) records...since that would break the humdrum syntax.)


# General Filtering



The indexing commands (previous section) only get you so far.
If you want to be more precise about filtering, use the `subset` nethod for `humdrumR` data.
`Subset.humdrumR` works exactly like [within.humdrumR](WorkingWithData.html "Working with Humdrum Data").
However, the "do" command you give must evaluate to a logical vector (`TRUE` or `FALSE`) of the same length as its input.
(You will get an error otherwise.)
`Subset.humdrumR` will take your logical result and filter out data which matches `FALSE`.

We can reproduce the functionality of the `[]` and `[[]]` indexing operators (previous section) using `subset`:


```{r}
subset(chorales, Spine == 1)

```

But, now we can filter based on *any* arbitrary criteria we want.
For example, we could extract tokens from odd numbered spines in odd numbered records *and* even numbered spines in even numbered records.
We'll use `R`s modulo command `%%` to separate even and odd numbers (`odd %% 2 == 1`, `even %% 2 == 0`).

```{r}

subset(chorales, (Record %% 2 == 0) == (Spine %% 2 == 0) )

```

Would we ever want to do that? Probably not.
However, returning to our flats, study, lets grab all the flat notes:

```{r}

subset(chorales, Token %grepl% '-')


```


Woh, we're only seeing a could of B flats in these files!
But from before, we know that there are a lot more flats in the fourth file (after losing the files with no flats):

```{r}

subset(chorales, Token %grepl% '-')[4]

```


## Subseting by Group


Since `subset.humdrumR` makes use of [within.humdrumR], we can use `within.humdrumR`'s other special options, including group `by`.
This can be a useful way to get some context for our searches.
For example, let's say we want to find flats again, but we want to see the whole *bar* of music that contains a flat.
We can do this by grouping by the `Bar` field (and the `File` field).
We'll want to say "within each bar, if there are *any* flats, return `TRUE` for the *whole* bar, else return `FALSE` for the *whole* bar."
We can use the `any` function to get *any* match, and the `dofill` option to copy the result of `any` to the whole bar.

```{r}

subset(chorales, dofill = any(Token %grepl% '-'), by = list(File, Bar))


```


Not enough context for you? maybe we should group the bars into even-odd pairs:

```{r}

subset(chorales, dofill = any(Token %grepl% '-'), by = list(File, floor(Bar  / 2)))

```


## Removing vs Filtering

You probably noticed that, unlike the indexing commands `[]` and `[[]]`, `subset` doesn't seem to actually *remove* data that you filter out.
So when we say 

```{r}
subset(chorales, Spine == 1)

```

we still have four spines, but spines 2--4 are just emptied.
This is correct.
What `subset` actually does is mark filtered data in the `Filter` field of the humdrum table.
`humdrumR` then ignores that data automatically.
Why do we do this?
There are several reasons:

1. Sometimes, seeing the full structure remain in place is cleaner/easier to interpret than actually removing them.
   Basically, all those empty spines remind you what you filtered, so it is easier to see if the filter is doing what you     *think* it is doing.
   + A clear example of this is when we did `[[ , j]]` indexing for flats (above): when we simply removed the spines with no flats it was hard to tell which spine was which in the result. 
2. Many filters would break the humdrum syntax if the data was simply removed. If you removed all the tokens that don't        contain flats, the result would be humdrum data with holes in it.
3. Its possible to undo your filters, using the `clearFilter` command. (Try it.) 

Once you've done some filtering with `subset`, if you *want* to get rid of empty parts of the data, you can do so using the commands `removeEmptyFiles`, `removeEmptySpines`, `removeEmptyRecords`, `removeEmptyStops`, or `removeEmptyPaths`.
By using this commands, we make sure that 1) you explicitely want to remove them and 2) the humdrum syntax is not broken, because whole records/spines/paths/files are removed.

So for our spines example:

```{r}

subset(chorales, Spine == 1) |> removeEmptySpines()

```

or for records:

```{r}

subset(chorales, Record %% 2 == 0) |> removeEmptyRecords()


```

