---
title: "Counts, Probability, and Information"
author: "Nathaniel Condit-Schultz"
date:   "August 2024"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Counts, Probability, and Information}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: sentence
---

```{r, include = FALSE, message=FALSE, echo = FALSE }
source('vignette_header.R')
```


Welcome to "Counts, Probability, and Information"!
This article explains some of the features R and  `r hm` have for extracting count distributions, probability distributions, and information theory metrics from data.


# Counting things

Nearly all of our articles, from the very [start](GettingStarted.html#counting-things "Getting started with humdrumR > Counting things"), have made use of the `count()` function.
This is a `r hm` workhorse function, which is used to count the number of unique values in a singe vector, or unique *combinations of values* across multiple of vectors.


Let's take a moment to explain all of what `count()` can do, starting from the basics.
The `count()` function accepts one or more atomic vectors, all of which must be the same length.
It counts all the unique combination of values across the vectors, and builds a table.
This can be illustrates with some simple examples:

```{r}

v1 <- c('a', 'b', 'c', 'a', 'b', 'c', 'a', 'b', 'c', 'a')
v2 <- c('A', 'A', 'A', 'B', 'B', 'C', 'C', 'C', 'C', 'C')


count(v1)
count(v2)
count(v1, v2)


```

Each input vector is considered one data dimension.
You can name your dimensions by naming the input arguments:

```{r}
count(lowercase = v1, uppercase = v2)

```

If there are `NA` values in your data, they will be counted unless you specify `na.rm = TRUE`.

```{r}

v1[3] <- NA

count(v1, v2)

count(v1, v2, na.rm = TRUE)

```

There is also an option to sort the table. 
You can use `-1` for increasing sort.

```{r}
count(v1, v2, sort = TRUE)
count(v1, v2, sort = -1)
```




#### Counting Real Numbers 

If you have a distribution of continuous numbers, you probably don't want to count every unique value (because there will be one of each).
Thus, `count()` will (attempt) to treat numeric values in a smart way by binning them into ranges using the base R `hist()` function's binning algorithm.
So for example:

```{r}
numbers <- rnorm(1000, 2, 2)

numbers |> head(30)

count(numbers)

```


The binning process can be controlled using the `binArgs` argument, which must be a list of arguments passed to [hist()].
A particularly useful one is `breaks`, which controls the number of, or locations of, bins.

```{r}

count(numbers, binArgs = list(breaks = 12))
count(numbers, binArgs = list(breaks = -8:12))

```

----

If you *don't* want to bin numbers, just apply `as.character()` (perhaps after rounding the numbers).

```{r}

numbers |> round(1) |> as.character() |> count()

```


> In base R, the `table()` function serves basically the same role of the `count()` function.
> You can use `table()` in many areas where you can use `count()`---you'll see that their output is essentially equivalent, but in a different structure (a `matrix` rather than a `data.frame`).
> However, `count()` is fully integrated into the `dplyr` universe and workflow, so we recommend using it.


### Counting real data


Enough with the artificial examples; let's have a go at using [count()] on some musical data.
This should feel familiar, given we've done similar stuff before in other articles:

```{r}

bach <- readHumdrum(humdrumRroot, 'HumdrumData/BachChorales/.*krn')

bach |> 
  mutate(Pitch = kern(Token, simple = TRUE),
         Rhythm = recip(Token)) -> bach

bach |> count(Pitch, na.rm = TRUE)

bach |> count(Rhythm)

bach |> count(Pitch, Rhythm, sort = TRUE)

```

Woh, there are a lot of `.` values in that last table, which is a shorthand for "zero."
This is because [count()], by default, recognizes every possible combination of values that could occur, even if they never occur.
Do get rid of these unused levels, we can set `.drop = TRUE`.

```{r}
bach |> count(Pitch, Rhythm, sort = TRUE, .drop = TRUE)

```


However, we might find all those `1`s uninteresting as well.
Luckily, the output of `count()` is a data.frame, and we can filter it:

```{r}

bach |> 
  count(Pitch, Rhythm, sort = TRUE) |>
  filter(n > 20)

```

Or maybe we'd like to see only the counts for quarter and eighth notes:

```{r}

bach |> 
  count(Pitch, Rhythm, sort = TRUE) |>
  filter(Rhythm %in% c('4', '8'))

```


## Probabilities

A close partner to [count()] is the [pdist()] function, which computes the *proportion* of input data is each unique value---which we can interpret as *probabilities*:


```{r}

pdist(v1)
pdist(v2)

pdist(v1, v2)


```

Pdist has all the same arguments as [count()] (`sort`, `.drop`, etc.), plus one other important argument: `condition`.
This argument can be used to calculate the *conditional probabilities*.
You can indicate one dimension to condition on, either by index of name:

```{r}

pdist(v1, v2, condition = 2)

pdist(lower = v1, upper = v2, condition = 'lower')

bach |>
  pdist(Pitch, Rhythm, condition = 'Rhythm')

```

So here we are seeing the probability of each pitch value *given* each possible rhythmic value.

## Marginal disributions

When you have a multi-dimensional count/proportion table, you might want to "collapse" one dimension to get what's called the marginal distribution.
You can do this with a [count()]/[pdist()] table by simply indexing by dimension (either by name or index):

```{r}

bach |>
  count(Pitch, Rhythm) -> twodim

bach |> 
  pdist(Pitch, Rhythm, Spine) -> threedim


twodim[ , 'Rhythm']

threedim[ , 'Rhythm']
threedim[ , 2:3]

```


## Combining tables

Sometimes, we make multiple tables of different data samples:



```{r}


chor1 <- bach |> filter(File == 1) |> kern(simple = TRUE) |> count()
chor2 <- bach |> filter(File == 2) |> kern(simple = TRUE) |> count()

chor1
chor2

```

Notice that our `chor1` and `chor2` tables have slightly different categories---for example, the second chorale includes D#s, G#s and A#s which never occur in the first chorale.
What if we want to compare, or combine these two tables?
Luckily, `r hm` makes this easy:

```{r}
chor1 + chor2

cbind(chor1, chor2)

```

Note that this won't work unless the tables have the same dimension names---it worked here because both tables have one dimension called `Kern`.
Also, notice that the result of using `cbind` is now a "plain" R data.frame---this is because it is not a single distribution of data anymore, but rather, two distriubutions pasted together.

# Modeling probability and likelihood


Statisticians refer to the sorts of distributions created by [count()] or and [pdist()] as *empirical distributions*---distributions of a sample of data.
We will often regard an empirical distribution as an estimate, or *model*, of some real-world probability distribution.
For example, the Bach chorles are often considered an idealized model of "common practice tonality" in general.
If we observe, for example, that $\approx$1.04% of notes in our Bach chorale data are E flats, we can (if would be so bold) that aboud 1.04% of notes in "common practice music" more generally are E flats.
Or to put it in a probabilistic way, if we randomly select a note from a common-practice piece, the probability of that note being an E flat is about 1.04%---according to our model, which is based on our empirical analysis of the first 10 chorales.


### Likelihood

Sometimes, we'd like to ask what probability a model assign to each data point.
This kind of "retrospective probability" assignment is called a *likelihood*.
In `r hm`, we can assign likelihoods quite easily using the [like()] function.
Consider this simple vector again:

```{r}

v1 <- c('a', 'b', 'c', 'a', 'b', 'c', 'a', 'b', 'c', 'a')

p1 <- pdist(v1)

```

If we imagine that `v1` is created by randomly drawing from an underlying probability distribution, then `p1` is an estimate of that probability distribution---`p1` is our "model" of `v1`.
We can then compute the estimated likelihood of observing each data point, given that model:

```{r}

like(v1)

```

What's happening here?
Notice that [like()] assigns each data point a likelihood by looking that value up in the `p1` table.
For example, `b` makes up `.3` of `v1`, so every time there is `b` in the input, [like()] assigns `.3`.
If we do this with our Bach data, we'd see that every E flat that appears gets that 1.04% value that we just talked about:


```{r}

bach |>
  like(Pitch)

```


We can extend this to two (or more) dimensions, where each likelihood is determined as th empirical joint probability of the input variables:

```{r}

bach |> 
  like(Pitch, Rhythm)

```


Here's a thought: what if we know the rhythm value in advance, and we want to estimate the likelihood of each pitch given the rhythm.
This is the "conditional likelihood," and we already say how we can do this with [pdist()]:

```{r}

bach |> 
  within(like(Pitch, Rhythm, condition = 'Rhythm'))

```

### Using a different model

Remember, there is no "true" or correct probability model for a given data.
We should never say the "probability of E flat *is* 1.04%"---rather, the "estimated probability of E flat."
It is very common, as a basic analysis step, to compute the empirical distribution of a data sample, then immediately use that distribution as a probability model of that data...essentially, using the data as a model of itself.
This is what [like()] does by default.
However, we might sometimes want to take a different tack, assigning likelihoods to our data based on a different model.
For example, what if we observed a different sample of lowercase letters:

```{r}
v3 <- c('a', 'a','b','b','b','b','b','c','c')
```

There's a much bigger proportion of bs here then we saw in `v1`.
If we, for whatever reason, think the `v1` model we had is a better model of the "true" probability distribution we are interested in, we could assign likelihoods to `v3` but use the `v1` probability model.
To do this, use the `model` argument.
Look what the difference is:


```{r}

like(lower = v3)

like(lower = v3, model = pdist(lower = v1))

```

We do this sort of thing with real data all the time.
For example, we might want to assign probabilities to the notes in the first Bach chorale, based on the empirical probabilities observed in all the *other* chorales?

```{r}


bach |>
   filter(File != 1) |>
   kern(simple = TRUE)  |>
   pdist() -> allExcept1

bach |>
  kern(simple = TRUE) |>
  within(like(Kern, model = allExcept1))
```



# Information Theory



Working with actually likelihoods, as we do above, is rarely a good idea, because in practice the likelihoods can get so small that we run into problems representing them in our computers.
Thus, we usually, compute the log of the likelihood, like this:

```{r}
bach |>
  kern(simple = TRUE) |>
  within(log(like(Kern, model = allExcept1)))
   

```

That's the "log likelihood".

Log likelihoods are always negative numbers, with more negative numbers indicating lower probability.
By convention, we like to invert that relationship by multiplying the log likelihood by `-1`.
If we do this, we get whats called the *information content* of each data point.
The smaller the information content the *more likely*---and thus more predictable and less informative---a data point is.
The higher the information content, the *less likely*---and thus more surprising and more informative---a data point is.
This way of interpreting inverted log probabilities is the core of **information theory**.
We can estimate the information content directly using the [info()] function, which is used just like the [like()] function:

```{r}

info(v1)

bach |>
  kern(simple = TRUE) |>
  within(info(Kern, model = allExcept1))
```


## Entropy

Information content is related to the (log)likelihood, in that we are assigning a value to every data point.
If we have a random variable (or sample from one), we can also ask how much information each data point has on average.
This is called the (Shannon) **entropy**.
Entropy is simply the average information content---which is really a property of the probability distribution.

```{r}

v1 |> pdist() |> entropy()


bach |>
  kern(simple = TRUE) |>
  entropy()

```

We should remind ourselves that what we are computing here is the *empirical entropy*---the entropy of the probability distribution we are estimating from our data.
The *true* entropy of pitch in common practice music is certainly not this exact number, but we might argue/hope that the empirical entropy we observe in these ten Bach chorales is a reasonable estimate of the true entropy of common practice music.

### Multi-dimensional entropy

The idea of entropy can be extended to multiple dimensions---its really no different than what we've been doing throughout this article.
We can compute the empirical *joint* entropy of two or more variables like this:


```{r}

bach |>
  entropy(Pitch, Rhythm)

bach |>
  entropy(Pitch, Rhythm, Spine)

```


Alternatively, we can compute the *conditional entropy*---which is simply the average information content of the *conditional probabilities*.
As before, we just need to provide that `conditional` argument:

```{r}
bach |>
  entropy(Pitch, Rhythm, condition = 'Rhythm')

bach |> with(entropy(Pitch, Pitch[lag = -1], condition=2))

```


### Mutual Information




### Cross entropy

The entropy of the conditional probability distribution is called the conditional entropy.

## Mutual Information

### Using a different model

If we compute entropy of one dataset using a different probability mode, this is called the cross entropy:

```{r}

bach |>
  kern(simple = TRUE) |>
  with(xentropy(Kern, model = allExcept1))
```

Notice that the cross entropy is always going to be higher than the entropy, because the entropy is estimated directly on itself, it "fits" its distribution as well as it can.
The difference between the cross entropy and the entropy is called the [Kullback-Leibler Divergence](https://en.wikipedia.org/wiki/Kullback%E2%80%93Leibler_divergence), more commonly *kld*.
If the kld is small, it means the model matches the data pretty closely.

