---
title: Rhythm and Pitch Intervals
author: "Nathaniel Condit-Schultz"
date:   "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Rhythm and Pitch Intervals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

`humdrumR` contains a rich set of built in representations and functions for representing musical tonality and meter.
A core philosophical/design choice is that both structures are conceptualized/encoded as *intervals*.



# Intervals 

Both pitch and time are perceived *relatively*: as a relationship between two points.
Thus we represent them, conceptually, as an *interval*---a difference.
In the case of pitch, an interval between two frequencies (or more abstractly, a "tonal" interval on the circle-of-fifths).
In the case of rhythm, an interval between two time points---between two onsets, or between an onset and an offset.
In `humdrumR`, these abstract intervals are represented as `tonalInterval` and `rhythmInterval` objects, respectively.

To concretize our abstract intervals, we must establish what they are relative to. 
What is the reference?
There are actually multiple, useful ways we can represent the same information.
Since all intervals are relative, there is always an *implicit* reference---the origin, zero.
For `tonalIntervals` the origin is the unison interval.
For `rhythmIntervals` the origin is zero.

`tonalInterval`s and `rhythmIntervals` constitute each constitute an algebraic *module over integers* (module over $\mathbb{Z}$), with appropriate arithmetic operations fully defined in `humdrumR`.
Since intervals can be added/subtracted, any interval can be thought of as a combination (addition) of other intervals.
In `humdrumR`, we often "partition" a `interval` as a combination of useful sub-intervals.

## Serial and Fixed Reference

Vectors of `humdrumR` intervals can be represented in two fundamental ways: *serial reference representation* and *fixed reference representation*.
To illustrate, we will use two examples, one rhythmic, one pitched:

1. The major scale
2. The "tag-line" rhythm `3+3+3+3+2+2`


### Fixed reference

In a fixed-reference representation, all intervals are interpreted relative to a fixed point (implicitly 0).
Fixed-reference representations of the major scale (using semitones) and the tag-line rhythm are as follows:


```{r}

referenceMajor   <- c(0, 2, 4, 5, 7, 9, 11, 12)

referenceTagline <- c(0, 3, 6, 9, 12, 14, 16)

```

Note that this representation is quite commonly used by musicians, in various forms, *for pitch*.
In contrast, fixed-reference representations are **not** commonly used by musicians *for rhythm*.
However, they *are* used in some contexts, as they essentially represent "time from the beginning" of a piece:
Thus, the timestamps/timelines in DAWs or "offsets" in `music21` are examples of fixed-reference representations of time.

### Serial reference

An alternative representation is a serial-reference representation, where each interval is measured relative to the previous one.
Thus, each data point represents the local change ($\Delta$) in the parameter.
To be lossless (as explained below) the first element of a serial data vector is measured relative to the implicit reference (0).
Serial representations of our tag-line and the major scale are as follows:

```{r}

serialMajor   <- c(0, 2, 2, 1, 2, 2, 2, 1)

serialTagline <- c(0, 3, 3, 3, 3, 2, 2) 

```

This representations is also fairly intuitive to musicians---representing "melodic intervals"---but is *not* how music notation or sequencers represent pitch.
In contrast, the serial-reference representation for rhythm *is* normative: traditional music notation note values are a serial-reference representation.


### Transformations ("Interval Calculus")


Since addition is defined for `interval`s fixed-reference and serial-reference representations can be translated between each other.

A serial-reference representation can be calculated as the pairwise differences between elements in a fixed-reference data, with the first reference element appended at the beginning.
Conversely, a fixed-reference representation can be calculated as the cumulative sum of serial-reference data.
Thus:

```{r}

identical(cumsum(serialMajor), referenceMajor)

identical(cumsum(serialTagline), referenceTagline)

identical(c(referenceMajor[1], diff(referenceMajor)), serialMajor)

identical(c(referenceTagline[1], diff(referenceTagline)), serialTagline)


```

In `humdrumR`, we refer to these two transformations as `delta` ($\Delta$) and `sigma` ($\Sigma$), for differences and sums respecticely.
In `humdrumR`, we note that the relationship between fixed- and serial-reference representations is analogous to the relationship between a function and its derivative.
The fixed-reference representation represents a sequence of independent values as a function of (serial) index position.
The serial-reference representation represents the same sequence as the differences between adjacent elements at each index---how much the fixed representation *changes* at each index.
(Since the first element in a serial-reference representation is relative to a fixed (implicit) reference, the normal one-to-many relationship between derivatives and functions is removed, and a lossless one-to-one relationship is maintained.)
These two transformations are thus inverses of each other: $x = sigma(delta(x))$ and $x = delta(sigma(x))$.
<!--These two transformations are thus inverses of each other: $x = integrate(derive(x))$ and $x = derive(integrate(x))$.-->

<!--Is is because of this analogy that `humdrumR` refers to these functions as `derive` (pairwise differences) and `integrate` (cumulative sum) respectively.-->


```{r}
library(humdrumR)

identical(referenceMajor, sigma(delta(referenceMajor)))

identical(referenceMajor, delta(sigma(referenceMajor)))


```


```{r, fig.cap = "Derivative/Integral relationship in Major Scale", echo = FALSE, results = "hide"}
library(humdrumR)

plot(c(0, referenceMajor[-1] - .25), type = 'h', xlab = 'Index', ylab = 'Semitones',
	col = 'blue', main = "")
text(1:8, referenceMajor, labels = referenceMajor, col = 'blue', cex=.7)

points(referenceMajor, type = 'c', col = 'red')
text((2:8) - .5, referenceMajor[-1] - (serialMajor[-1]/2), labels = paste0(c('-', '', '+')[2 + sign(serialMajor[-1])], serialMajor[-1]), col = 'red', cex=.5, pos=1)

legend('topleft', fill = c('blue', 'red'), legend = c('Reference', 'Serial'), bty = "n")

```
```{r, fig.cap = "Derivative/Integral relationship in Tagline Rhythm", echo = FALSE, results = "hide"}
library(humdrumR)

plot(c(0, referenceTagline[-1] - .25), type = 'h', xlab = 'Index', ylab = 'Ticks',
	col = 'blue', main = "")
text(1:8, referenceTagline, labels = referenceTagline, col = 'blue', cex=.7)

points(referenceTagline, type = 'c', col = 'red')
text((2:8) - .5, referenceTagline[-1] - (serialTagline[-1]/2), labels = paste0(c('-', '', '+')[2 + sign(serialTagline[-1])], serialTagline[-1]), col = 'red', cex=.5, pos=1)

legend('topleft', fill = c('blue', 'red'), legend = c('Reference', 'Serial'), bty = "n")

```





## Dynamic Reference

Since `intervals` are additive, another possibilty is to use one vector of intervals as the reference point for a second vector of intervals.
This allows us to create dynamic reference points.

Some examples 

+ **Figured Bass**: the bass voice in the texture is encoded (in serial/reference reprsentation) while other voices are represented relative to the bass voice.
+ **Chord Tones**: the abstract "root" of chord progression is represented in one vector, with other vectors representing the "chord tones" (root, 3rd, 5th, 7th, etc.) relative to chord root.
+ **Scale degrees**: The tonic of the key is used as a dynamic reference, with notes calculated relative to the tonic.

```{r, fig.cap = "Figured Bass", echo = FALSE, results = "hide"}

library(humdrumR)

# In semitones:

bassline_reference <- c(0, 7, 4, 5, 2, 4, 5, 7, 7)

harmonicintervals <- c(16, 7, 8, 4, 15, 12, 9, 5, 4)

plot(c(0, bassline_reference[-1] - .25), type = 'h', xlab = 'Index', ylab = 'Pitch', ylim = c(0, 19), axes = FALSE,
	col = 'blue', main = "")
text(seq_along(bassline_reference), bassline_reference, labels = bassline_reference, col = 'blue', cex=.7)

ax <- c(0, 2, 4, 5, 7, 9, 11, 12, 14, 16, 17, 19)
# axis(2, ax, as.tonalChroma(as.character(ax)), tick = FALSE, las = 1)

i <- seq_along(bassline_reference)

	segments(x0 = i, x1 = i, y0 = bassline_reference + .25, y1 = bassline_reference + harmonicintervals - .25, col = 'red')
	text(i, bassline_reference + harmonicintervals, labels = harmonicintervals, col = 'red', cex = .7)


legend('topleft', fill = c('blue', 'red'), legend = c('Bass Interval', 'Harmonic Interval'), bty = "n")

```

Calculating dynamic reference intervals is lossless only if the complete information on the reference intervals are maintained.
For instance, we we keep track of the serial or fixed intervals off the bass voice.
Or, if we keep track of the local key (usually using a fixed interval to "C").


