---
title: "Overview of humdrumR"
author: "Nathaniel Condit-Schultz"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Overview of humdrumR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}

hm <- "humdrum$_{\\mathbb{R}}$"
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Welcome to "Overview of humdrumR"!
This document will give you a very brief overview of what the `r hm` package is, and what it can do. 
If you're all ready to start learning how to use `r hm`, checkout [getting started with humdrumR](GettingStarted.html "Getting started with humdrumR")

# What is `r hm`?

`r hm` is is a R package for the visualization, manipulation, and analysis of data encoded in the Humdrum syntax [http://www.humdrum.org].

The name of the package is pronounced _hum-drum-**ARRRRR**_, with the last syllable emphasized in pirate fashion.
The name should always be typeset $h\mu m_{\mathbb{R}}^{\Delta r \mu m}$.
Failure to pronounce or typeset $h\mu m_{\mathbb{R}}^{\Delta r \mu m}$ correctly will void the warranty.
 

## What is humdrum?

Humdrum is a system for analyzing blah blah.

```{r setup}
library(humdrumR)
```


