---
title: Pitch and tonality in humdrumR
author: "Nathaniel Condit-Schultz"
date:   "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pitch and tonality in humdrumR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE, message=FALSE, echo = FALSE }
source('vignette_header.R')

par(mar=c(1,1,1,1))
pitchFunctions <- humdrumR:::pitchFunctions
```


As a computational musicology toolkit, `r hm`'s tools for analyzing and manipulating pitch data are just about the most important tools in the toolbox.
For the most part, `r hm`'s pitch tools are focused on the Western system of tonality, and tools for representing pitches in a tonal context are the focus of *this* vignette---we also have standard tools for looking at pitch from a Western, 12-tone atonal setting as well.


# Pitches and Intervals


`r Hm` defines a suite of "pitch functions," like `kern()`, `solfa()`, `interval()`, and `semits()`.
These functions all work in essentially the same way: they take in an input argument and output pitch information in their own particular format.
For example, let's take a little bit of `**kern` data:

```{r}

input <- c('4.c', '8d', '4e', '2.g')

```

Notice that this data has both rhythmic information (`4.`, `8`, `4`, `2.`) *and* pitch information (`c`, `d`, `e`, `g`).
What happens if we give this data to our "pitch functions"?:

```{r}

kern(input)
interval(input)
semits(input)
solfa(input)

```


Each function correctly reads the `**kern` pitch information, ignoring the rhythm information, and outputs the pitch information in a different format.
This allows us to "translate" between different ways of representing pitch information.

If you *want* to keep the non-pitch (rhythm) information, use the `inPlace` argument:

```{r}

kern(input, inPlace = TRUE)
interval(input, inPlace = TRUE)
semits(input, inPlace = TRUE)
solfa(input, inPlace = TRUE)

```


The cool thing is that each of these functions can read any of the other functions' output.
So you can do things like:

```{r}


kern('Cb6')

pitch("eee-")

pitch(-4) # semits

kern('-4')

solfa('A#6')

semits('Ab3')



```


----

The complete list of basic pitch functions is: 

```{r results = 'asis', echo = FALSE}

pfs <- rapply(humdrumR:::pitchFunctions, 
                \(func) paste0('    + `', 
                                ifelse(humdrumR:::.names(func) == '', func, paste0(humdrumR:::.names(func))), 
                                '`', ifelse(humdrumR:::.names(func) == '', '', paste0(' (', func, ')'))), how = 'list')

pfs <- lapply(pfs, \(top) Map(\(name, pf) paste(c(paste0('  + *', name, ' pitch representations*'), pf), collapse = '\n'), names(top), top))

pfs <- Map(\(name, l) paste(c(paste0('+ **', name, ' pitch representations**'), unlist(l)), collapse ='\n'), names(pfs), pfs)
cat(unlist(pfs), sep = '\n')


```

Each one of these functions represents a different way of representing equivalent pitch information.


#### Documentation

The global documentation for *all* the pitch functions can be seen by calling `?pitchFunctions`.
You can also call documentation for any individual function, like `?kern`.

## Scale degree

Some pitch representations encode [scale degree](https://en.wikipedia.org/wiki/Degree_(music)), which depends on the key.
We can use the `Key` argument to control how our pitch functions interpret key---either translating *from* degrees to absolute pitches, or vice versa.
Pass the `Key` argument a humdrum key interpretation, like `c:` (c minor) or `A-:` (a-flat major).

```{r}

absoluteInput <- c('c', 'd-', 'e-', 'g', 'a-')
solfa(absoluteInput)
solfa(absoluteInput, Key = 'A-:')

degreeInput <- c('do', 're', 'mi', 'fi', 'so')

kern(degreeInput)
kern(degreeInput, Key = 'A-:')

```

----

The `Key` argument is vectorized (see the [R primer](RPrimer.html) if you don't know what that means!); that means that you can actually read information with different/changing keys.



## Working with pitch representations

Let's have a look at using pitch functions on real data.
Let's load `r hm`'s built-in Bach chorales:

```{r}
chorales <- readHumdrum(humdrumRroot, 'HumdrumData/BachChorales/.*krn')

```
These chorales are full of `**kern` data in the (default) `Token` field, which can easily be parsed/translated.
Maybe we'd like to convert the `**kern` to semitones.
`r Hm`'s pitch functions can all be used "[humdrum-style](DataFields.html#humdrum-style "HumdrumR data fields article, humdrum-style section"), so using `semits()` is very easy:

```{r}

chorales |>
  semits()

```

We could then, for instance, make a histogram of the semitone values:

```{r}

chorales |>
  semits() |>
  pull() |>
  hist(main = 'Histogram of semits() in 10 chorales')

```

### Keys

The chorale dataset has key information in its built-in `Key` field.
If we use a pitch function in humdrum- or -tidy style, the `Key` field is *automatically* passed as a `Key` argument to the function!
That means if we apply `solfa()`, we will get the correct scale degrees, *given* the keys in the data:

```{r}

chorales |>
  solfa(Token)

```

## Rests

`**kern` data, like in these chorales, includes *rests*, indicated by `r` in place of a pitch.
What happens when a pitch function sees a rest in our data?
Let's see:

```{r}

solfa(c('8gg', '8ff', '4ee', '8dd', '8r', '2cc'))

```

The rest token, is output as a null (`.`) value---this null `.` is actually an R `NA` value.
In fact, any token in the input which fails to parse as pitch will return as `NA`/`.`.
For example:

```{r}
kern(c('C4', 'X', 'E4'))

```


# Pitch arguments

Since all the "pitch functions" do the same sort of task, they share a number of common arguments.
The complete list of these pitch arguments can be found in the `?pitchParsing` and `?pitchDeparsing` man pages, but the most important ones are described here:


#### Generic vs Specific pitch

All pitch functions accept logical (`TRUE` or `FALSE`) `generic` and `specific` arguments.
If `generic == TRUE`, only generic pitch information is returned.
This affects different pitch representations in different ways:

+ For representations like `kern()`, `pitch()` and `lilypond()`, no accidentals are included in generic output;
+ For `intervals()`, the interval quality is removed;
+ For `solfa`, scale degree four will be "fa", never "fi" (sharp four).

The `specific` argument is simply the opposite of `generic`, so you have the choice of indicating `genric = TRUE` or `specific = FALSE`---they are equivalent.

---

In code:

```{r}

input <- c('E', 'G#', 'A', 'B-', 'F#', 'G')

pitch(input, generic = FALSE)
pitch(input, generic = TRUE)

interval(input, specific = TRUE)
interval(input, generic = TRUE)

solfa(input, specific = TRUE)
solfa(input, specific = FALSE)


```

#### Simple vs Compound pitch

All pitch functions accept logical (`TRUE` or `FALSE`) `simple` and `compound` arguments.
If `simple == TRUE`, octave information is stripped from the output, leaving only "simple" pitch information.
This affects different pitch representations in different ways:

+ For `pitch()` and `degree()`, the octave number is removed.
+ For `lilypond()` and `helmholtz()`, the octave marks `'` and `,` are removed.
+ For `solfa()` and `deg()`, the contour indicators (explanation below) `v` and `^` are removed.
+ For `interval()`, only steps between 1--7 and returned (i.e., no 10ths).


The `compound` argument is simply the opposite of `simple`, so you have the choice of indicating `simple = TRUE` or `compound = FALSE`---they are equivalent.

---

In code:

```{r}

input <- c('c', 'a', 'dd#', 'ee', 'G', 'GG', 'GG-', 'CCC')

degree(input, simple = FALSE)
degree(input, simple = TRUE)


lilypond(input, compound = TRUE)
lilypond(input, simple = TRUE)

semits(input, compound = TRUE) 
semits(input, compound = FALSE)

solfa(input, compound = TRUE)
solfa(input, compound = FALSE)

interval(input, simple = FALSE)
interval(input, simple = TRUE)
```

### Working with pitch arguments

The `generic` and `simple` arguments are useful if you want to tabulate pitch classes, for example:

```{r}
chorales <- readHumdrum(humdrumRroot, 'HumdrumData/BachChorales/.*krn')


chorales |>
  kern(simple = TRUE) |>
  tally() |>
  draw()

```

```{r}
chorales |>
  kern(generic = TRUE) |>
  tally() |>
  draw()
```

```{r}

chorales |>
  kern(simple = TRUE, generic = TRUE) |>
  tally() |>
  draw()
```


## Transposition

All of `r hm`'s pitch functions have built-in transposition functionality.
You can use this functionality by passing a named list of arguments to the `transpose()` function, as the `transposeArgs` argument to the pitch function call, like in this example using the `by` argument:

```{r}

kern(c('c', 'd', 'e', 'f', 'f#', 'g'), transposeArgs = list(by = 'M2'))

```

Another option is to transpose by key.
You can the `to` argument through to `tranpose()`.
Let's say we have music in A major that we'd like to transpose *to* C major:

```{r}
Amelody <- c('8A', '8B', '4.c#', '16d', '8c#','8f#', '8e', '8d#', '4e')

kern(Amelody, Key = 'A:', transposeArgs = list(to = 'C:'))

```

So the `Key` argument to `kern()` indicates the key the music is coming from, and the `to` argument in `tranposeArgs` indicates where to transpose it to.
Since humdrum/tidy functions automatically use `Key` information from a data set, this makes it very easy to, say, transpose all `**kern` tokens in your data to C major:

```{r}
chorales <- readHumdrum(humdrumRroot, 'HumdrumData/BachChorales/.*krn')

chorales |>
  kern(simple = TRUE, transposeArgs = list(to = 'C:')) |>
  tally() |>
  draw()

```

This looks exactly like calling `solfa()`:


```{r}
chorales |>
  solfa(simple = TRUE) |>
  tally() |>
  draw()

```





# Melody and Harmony 

## Melodic Intervals

What if we want to calculate melodic intervals in each part of a score?
This the purpose of the `mint()` (melodic intervals) command.
We can use `mint()` on any pitch data:

```{r}

input <- c('d', 'B-', 'A', 'G', 'D', 'd', 'B-', 'G#', 'A', 'f', 'e', 'd')

mint(input)

```

If you use `mint()` in humdrum- or tidy- style on a `r hm` dataobject, `mint()` will automatically be called within each spine/path in each file (this is accomplished using the `groupby` argument to `mint()`):

```{r}

chorales <- readHumdrum(humdrumRroot, 'HumdrumData/BachChorales/.*krn')

chorales |> 
  mint()

```
`mint()` has a number of special options, which you can read about in the manual by checking out `?mint`.
For now, we'll just show you the `classify` argument, which can be used to classify output intervals as either `Unison`, `Step`, `Skip`, or `Leap`:

```{r}
chorales |>
  mint(classify = TRUE)

```



## Harmonic intervals

In parallel to `mint()` is `hint()` (harmonic intervals) command.
`hint()` works just like `mint()` except it calculates intervals from left to right in each record of a file:

```{r}
chorales |>
  hint()

```

A special trick is to use the `lag` argument to calculate harmonic intervals all in relation to the same spine.
For example, you can calculate harmonic intervals between each voice and the bass voice like so:

```{r}

chorales |>
  hint(lag = Spine == 1)

```
Or, harmonic intervals from the soprano like so:

```{r}
chorales |>
  hint(lag = Spine == 4)

```


Check out the "Logical lags" section of the `?hint` manual for details.