---
title: "Making plots with humdrumR"
author: "Nathaniel Condit-Schultz"
date:   "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Making plots with humdrumR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
source('vignette_header.R')
```

Welcome to "Making plots with `r hm`"!
R has two major plotting systems that are widely used: 1) R's built-in [base graphics][graphics-package] functions and 2) the more modern [ggplot2]. 
`r Hm` is compatible with both of these systems.
`r Hm` also has it's own, built-in [draw()] function, which is basically a spiffy (easier to use) extension of base-R graphics.

In this vignette we give detailed showcase of the features of [draw()].
We'll then show you how to use [ggplot2] with `r hm` data, if that floats your boat.
Finally, we go over some basic concepts from base-R graphics, which you can use if you want to customize [draw()] plots (or make your own custom plots from scratch).



-----


Let's load some humdrum data, and create some data fields to plot:


```{r}
bach <- readHumdrum(humdrumRroot, 'HumdrumData/BachChorales/.*hum')

```


We'll use [semits()] and [duration()] to convert pitch and rhythmic information into numbers, and [solfa()] and [recip()] to represent them in categorical form.

```{r}
bach |>
  mutate(Semits = semits(Token),
         Duration = duration(Token),
         Solfa = solfa(Token, simple = TRUE),
         Recip = recip(Token)) -> bach
```


# Draw()

`r Hm`'s [draw()] function is basically a fancy wrapper around base-R functions like [plot()] and [hist()].
Like these functions, using [draw()] is simple: just pass data ([atomic][vector] vectors) in to get a plot.
The [draw()] function will look at the input arguments you give it, and make an appropriate plot for that input type.
Currently, `draw()` can make seven different types of plots, depending on the input passed to it.

To decide what kind of plot to make, [draw()] looks at its first two arguments, `x` and `y`, which represent (you guessed it) the X and Y axes.
Here are the types of plots associated with different types of `x`/`y` arguments:

| `x`                                                      | `y`                                   | Plot type                                           |
|----------------------------------------------------------|---------------------------------------|-----------------------------------------------------|
| `numeric`                                                | (missing)                             | [Density Histogram/Contour][draw_density()]         |
| (missing)                                                | `numeric`                             | [Quantile plot][draw_Qplot()]                       |
|                                                          |                                       | (or Violin plot)                                    |
| `numeric`                                                | `numeric`                             | [Scatter/line plot][draw_scatter()]                 |                  
| `character`/`factor`                                     | `numeric`                             | [Violin plot][draw_violins]                         |
| `numeric`                                                | `character` or `factor`               | [Area chart][draw_area()]                           |
| `character`/ `factor`                                    | (missing)                             | [barplot][draw_barplot()]                           |
| (missing)                                                | `character`/`factor`                  | [barplot][draw_barplot()]                           |
| `character`/`factor`                                     | `character`/`factor`                  | [Heat map][draw_heat()]                             |


To see this in action, let's see what happens if we pass our `Semits` field as either the first (`x`) or second (`y`) argument to [draw()]:

```{r}

bach |> draw(Semits)

bach |> draw( , Semits)

```

On the X axis, numeric data is drawn as a [density histogram][draw_density()]; on the Y axis, a [quantile plot][draw_Qplot()].
What if we pass two numeric variables (`Semits` *and* `Duration`)? We get a scatter plot:

```{r}

bach |> draw(Duration, Semits)

```

However, there are only `r humdrumR:::num2word(length(unique(bach$Duration)))` duration values in the data, so maybe it makes more sense to treat them as categories, not numbers.
If we convert `Duration` to a `character`-string vector, see what changes:


```{r}

bach |> draw(Recip)

bach |> draw(Recip, Semits)

```
### a


# GGplot


```{r}
library(ggplot2)
```


```{r}

bach |>
  ggplot(aes(Duration, Semits)) + geom_point()

```


# Base-R 


The base-R graphics system is extremely powerful, giving you precise control of every aspect of plots.
Unfortunately, though *getting started* with simple base-R plots is extremely fast and easy, as we'll show you in a moment, it takes quite a while to learn on how to precisely control your plots, and it ends up being quite verbose.
The `r hm` [draw()] function is actually based on base-R graphics, and is fully compatible with base-R---our recommendation is to start with [draw()], then gradually learn base-R as needed.
(Alternatively, just learn [ggplot][#ggplot].)
If you are just getting started, you can probably skip ahead to the [Draw](#draw) or [GGplot](#ggplot) sections below, and come back to this for added context if and when you need it.


### Primary plot functions

Base-R graphics is centered around a few main plotting functions: [plot()], [barplot()], and [hist()].
Additional functions can then be used to *add* more elements on top of the plots you already created: [points()], [arrows()], [abline()], [lines()], [axis()], [polygon()], etc.
(These "adding" functions can be used with [draw()] too.)


We can then see what happens if we pass numeric data to [plot()] or [hist()].
(As usual, we have to use [with()] to access the fields.)

```{r}
bach |>
  with(hist(Duration))
```

```{r}
bach |>
  with(plot(Duration))

bach |>
  with(plot(Duration, Semits))


```

To use [barplot()], the standard approach is to tabulate categorical data using [table()], then pass that to `barplot()`.

```{r}
bach |>
  with(table(Solfa)) |>
  barplot()
```
### Adding plot functions

Once you've created a plot with [plot()], [barplot()], or [hist()], you can add more information on top of those plots using functions like [mtext()], [points()], or [abline()].

---

The [points()] function acts just like [plot()], except it doesn't create a new plot, instead adding the existing plot.
You could for example, use points to plot the average pitch value for each duration.
We will first remake the same plot as above using [plot()], then compute the averages add them as red points using [points()].


```{r}


bach |> 
  with(plot(Duration, semits())) 

bach |>
  group_by(Duration) |>
  summarize(MeanSemits = mean(Semits)) |>
  with(points(Duration, MeanSemits, col = 'red', pch = 16, cex = 1.2))
  


```

The arguments `cex`, `pch`, and `col` are used with [points()] (or [plot()]) to control the point size, point shape, and color respectively.



---

The [abline()] function can be used to straight lines on an existing plot.
You could, for example, draw a horizontal line separating the positive pitches from the negative pitches:



```{r}


bach |> 
  with(plot(Duration, Semits)) 

bach |> with(abline(h = 0))


```

You can also draw the linear regression line using [abline()], by passing it an [lm()] object:


```{r}


bach |> 
  with(plot(Duration, Semits)) 

bach |>
  with(lm(Semits ~ Semits)) |>
  abline()
  


```
