---
title: "Reading and writing data"
author: "Nathaniel Condit-Schultz"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reading and writing data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_knit$set(root.dir = humdrumR::humdrumRroot)
knitr::opts_chunk$set(message = FALSE,
  collapse = TRUE,
  comment = "#>"
)
hm <- "humdrum$_{\\mathbb{R}}$"
Hm <- "Humdrum$_{\\mathbb{R}}$"
```

Welcome to "Reading and writing data with humdrumR"!
This document will give you a detailed overview of `r hm`'s data reading and writing functionality.
This document closely parallels parts of `r hm`'s code documentation;
within an R session, you can read more concise version the material in this document by calling:

```{r } 
library(humdrumR)

?readHumdrum
?writeHumdrum

```

-------

> **Note**:
> `r Hm` is primarily intended to be used for analyzing existing humdrum data, and most users will be reading humdrum data a lot more often than writing it.
> `r Hm` *is* able to write modified humdrum data back to humdrum-syntax text files, but this is a secondary use-case, and has received relatively little attention/thought in the current version of the package.

# Reading humdrum data


The `readHumdrum` function is the gateway to `r hm`: it is the function we use to read humdrum data, encoded in humdrum-syntax text files, into R.

### Practice Data

In the following examples, we will make use the small number of raw humdrum data files which are included with `r hm`.
To access these files, we need to navigate to the directory where your computer installed `r hm`---fortunately, `r hm` records where this directory is on your computer in a variable called `humdrumRroot`.
Once we load the `r hm` library, we just need to set our R "working directory" to that location using the command `setwd`.

```{r}
library(humdrumR)
setwd(humdrumRroot)

```


The humdrum data is stored in a subdirectory called "HumdrumData"---we can look at the contents of this directory using `dir(recursive = TRUE)`:

```{r}

dir('HumdrumData', recursive = TRUE)

```

As we can see, there are `r humdrumR:::num2word(length(dir('HumdrumData')))` directories containing a total of `r humdrumR:::num2word(length(dir('HumdrumData', recursive = TRUE)))` files.


## Targeting files

To use `readHumdrum` we call the function with one (or more) character-string arguments.
(In the function documentation (`?readHumdrum`), you'll see these arguments called `...`---in R, `...` means *any* number of arguments.)
These arguments are interpreted as [regular expressions](https://en.wikipedia.org/wiki/Regular_expression), and matched against directories and files on your system.
We call these "**REpath-patterns**," short for Regular Expression directory-path patterns.
For example, 
```{r}
library(humdrumR)
setwd(humdrumRroot)

readHumdrum('HumdrumData/BachChorales/chor.*.krn')

```

uses the REpath-pattern `"HumdrumData/BachChorales/chor.*.krn"` to match all files in the 'HumdrumData/BachChorales' directory whose name matches the regular expression "chor.*.krn."
As you can see, the REpath-pattern includes directory (folder) paths (relative or absolute)---using your operating system's appopriate delimeter ("\" for windows, "/" for everybody else)---*and* a pattern for matching file paths.
*Each* directory is also treated as a regular expression, possibly matching multiple directories.
Let's break this down: `r hm` splits your REpath-pattern input string into three parts: `"HumdrumData"`, `"BachChorales"`, *and* `"chor.*.krn"`, and treats *each* of the three as regular expressions.
The the first two expressions are used to match directories, while the last expression is used to match files.
So, while `readHumdrum('HumdrumData/BeethovenVariations/.*.krn')` matches all the files with the ".krn" extension in the "BeethovenVariations" directories and `readHumdrum('HumdrumData/MozartVariations/.*.krn')` matches all the files in the "MozartVariations" directories, the command `readHumdrum('HumdrumData/.*Variations/.*.krn')` will match all kern files in both directories!


```{r}
readHumdrum('HumdrumData/.*Variations/.*.krn')
```

> Be careful: since each directory/file name is always treated like a regular expression, we can sometimes specify something that is more general than we intend. 
> For instance, the command `readHumdrum('MyFolder/.*')` will match files in a folder called "MyFolder," but would also match folders names "MyFolder_Also", or "ThisIsMyFolder."
> If you want to be sure to match exactly one and only one directory/file, use "^" and "\$" regular-expression markers to explicitely mark the beginning and end of your pattern:
> the command `readHumdrum('^MyFolder$/.*')` will only read files from one directory "MyFolder."

To read *all* of our test files we can enter:

```{r}

readHumdrum('HumdrumData/.*/.*')

```


> Note: Since `r hm` can read files from multiple directories at once, it is possible that two or more files with the same file name, but in different directories are matched.
> If this happens, `r hm` identifies each file with enough of it's directory path to make sure they are all unique.

----

### Multi-Argument Approach

Earlier we mentioned that we can have one *or more* of these RE-path patterns.
Why more than one?
Well, we can divided our RE-path patterns across multiple arguments (from left to right):
Instead of writing `readHumdrum("HumdrumData/BachChorales/chor.*.krn")`, we can write `readHumdrum("HumdrumData", "BachChorales", "chor.*.krn")`.
Thus these two approaches 

```{r eval = FALSE}
readHumdrum("HumdrumData/BachChorales/chor.*.krn")

readHumdrum("HumdrumData", "BachChorales", "chor.*.krn")
```

are identical.
What's the advantage of this approach? We'll see in the next section.

### Multiple Patterns

Sometimes, expressing all the files you want in a single regular expression is not possible.
Luckily, `readHumdrum` can accept as many separate patterns you want, just group them into a vector.
Instead of writing `readHumdrum('HumdrumData/.*Variations/.*.krn')`, you could be explicit and write 

```{r}
readHumdrum(c('HumdrumData/BeethovenVariations/.*.krn', 'HumdrumData/MozartVariations/.*.krn'))
```

I've used the `c` command to create a vector with two REpath-patterns.
However, that was a lot of (bug-prone) typing...what if we combine these multiple patterns with the Multi-Argument Approach from above?:

```{r}
readHumdrum('HumdrumData',
            c('BeethovenVariations', 'MozartVariations'),
            '.*.krn')
```
We get the same result!
How did this work exactly?
We've fed three arguments to `readHumdrum`.
The first and last arguments (`"HumdrumData"` and `".*.krn"`) are length one.
However, the middle argument (`"...Variations"`) is of length two.
`readHumdrum` concatenates the three arguments together, making *two* separate REpath-patterns:

```{r eval = FALSE}
"HumdrumData/BeethovenVariations/.*.krn"
```
**AND**
```{r eval = FALSE}
"HumdrumData/MozartVariations/.*.krn"
```


> Note:
> It is possible to write multiple patterns which match some (or all) of the same files. If the argument `multipleInstances = FALSE`, each unique file will only be read once (into the first matching pattern). 
> If `multipleInstances = TRUE`, the same file(s) can be read more than once.


#### Pattern Names

Whenever we specific more than one REpath-pattern, `r hm` gives them names which we can access in the `Label` field of the resulting `r hm` data object.
You can choose these labels by giving names to the patterns in your multi-pattern vectors:
Thus, we could write

```{r}
readHumdrum('HumdrumData',
            c(Rap  = 'Rap/.*',
              Rock = 'RollingStone/.*')) -> rapAndRock
```
and these two separate patterns will be matched and read *and* the names `"Rap"` and `"Rock"` will be associated with them in the resulting `Label` field.
This feature is very useful if you are working with multiple heterogeneous datasets and you want to be able to apply some analyses/parsing to only one of the data subsets.

### Contains

Sometimes you want to only read humdrum files which contain particular data.
For instance, you might want to read only the files in a minor key.
One option is to read all the files in the corpus then filter out the files you don't want (see the [humdrumR filtering vignette](Filtering.html) to learn how.
However, in some cases, it can save a lot of time and effort to filter the files you want before parsing them.
The `contains` argument of `readHumdrum` can be used in just this way!
The contains argument must be a character vector---each character string is treated as a regular expression, and only files which contain matches to all of these regular expressions will be read.
Thus, we could only read pieces with minor keys by writing:

```{r}
readHumdrum('HumdrumData', 'BachChorales', 'chor.*.krn',
            contains = '\\*[a-g][-b#]*:') 
```

(This regular expression matches standard humdrum tandem interpretations for minor keys.)

## Finding Files

`readHumdrum` interprets REpath-patterns and finds files using the helper function `findHumdrum`.
If you are not sure where your humdrum files are, or what pattern you want to use to find them, you might start be using `findHumdrum`.
`findHumdrum` takes the same input as `readHumdrum`, but doesn't fully parse the input into `r hm`.
Instead, it returns a `data.table` containing matching filenames (if any) and raw file content.

If you use the `verbose = TRUE` option for either `findHumdrum` or `readHumdrum`, a complete list of all the matches files will be printed.
This is useful to check if you are reading the files you *intend* to read.

### Validation

`findHumdrum`, and by extension `readHumdrum`, are smart functions that will ignore non-humdrum files matched by your pattern.
Non-text files (like `.pdf` or `.jpg`) will be ignored.
Whats more, text files which fail to conform to the humdrum syntax will not be read either.
You will see a message that checks which files are "valid humdrum," and tell you how many (if any) text files are not read.

If you are trying to read a file which you *think* should be valid humdrum, but `readHumdrum` won't read it, use `validateHumdrum` with the exact same REpath-pattern arguments, and specify and `errorReport.path` to see a detailed report of the problem.
`r Hm` won't read files with *any* humdrum-syntax violations---there might be minor problems in the data that you need to fix.
Call `?validateHumdrum` or read our [humdrum data validation vignette](Validation.html) to learn more about `r hm`'s data validation tools, and how they can be used to identify errors in your humdrum data.
(`validateHumdrum` does not automatically check if the humdrum *data* is encoded correctly or makes sense, just that it is structurally sound, in accordance with the humdrum syntax.)

## Reading

So, what actually happens when we run `readHumdrum`?
By now, you've seen that `readHumdrum` prints out a few messages describing its process.

1. First, the files which match your search patterns (if any) are identified, and their text is read into R.
2. Next, these files are checked to see if they are valid humdrum files. 
3. The files are parsed---i.e., translated into `r hm`'s [data.table backend](HumdrumTable.html). This step takes the longest by far, as it includes parsing all the humdrum files' data tokens, interpretations, reference records, and (worst of all) spine paths. Indeed, reading humdrum data with spine paths will take significantly longer than reading pathless data. To save yourself some time, you can potentially use the `tandems` and `reference` arguments (see "Parsing Metadata" below).


## Parsing Metadata

By default, `r hm` parses all metadata in the humdrum files it reads.
This includes the true global metadata for files, encoded in humdrum *reference records*, as well as the more local *tandem interpretations*.

### Reference Records

By default, `r hm` parses all reference records in the files it reads, creating a unique field for each reference key. 
(If there are more than one instance of any given key, they are combined in a string separated by semicolons.)
If there are a lot of unique reference keys in a dataset, they can end up taking a lot of memory in a humdrum table.
In these cases, you might find it useful to limit `readHumdrum` to only parsing those reference records which you are actually going to use.
This can be accomplished quite easily with the `reference` argument.

The `reference` argument defaults to `"all"`, which means all reference records are parsed.
However, reference can also be a character vector of reference codes.

Thus, 

```{r}
readHumdrum('HumdrumData',
            c('BeethovenVariations', 'MozartVariations'),
            '.*.krn',
            reference = 'COM')
```

will only parse the `COM` reference record in each piece.
We can use this to rename our reference fields if we like---in this case,

```{r}
readHumdrum('HumdrumData',
            c('BeethovenVariations', 'MozartVariations'),
            '.*.krn',
            reference = c(Composer = 'COM'))
```

the `COM` reference records will be parsed but saved into a field called `Composer` instead of `COM`.

If you don't want to parse *any* reference records, specify `reference = NULL`.


### Tandem Interpretations

Tandem interpretations are used in humdrum data to represent "local," real-time metainformation associated with specific spines---review our [humdrum syntax](HumdrumSyntax.html) vignette for more details.

`r Hm` can read any arbitrary tandem interpretations in humdrum data.
However, if non-standard interpretations are used, `r hm` has no way of knowing how to parse them.


By default, `r hm` always reads tandem interpretations into a field called `Tandem`.
This field tabulates all tandem interpretations that have so far appeared in a spine in order, concatenating them into a single comma-delimited, from most recent to most distant.
For instance, the `Tandem` field for the spine

```
**kern
*C:
*M4/4
*MM100
c e
c g
*D:
d f#
*-
```

is parsed as

```
""
C:
M4/4,C:
MM100,M4/4,C:
MM100,M4/4,C:
MM100,M4/4,C:
D:,MM100,M4/4,C:
D:,MM100,M4/4,C:
D:,MM100,M4/4,C:
```

So, the interpretations pile up into a longer and longer string as a piece progresses.
Notice that the `"D:"` and `"C:"` interpretations get piled up separately, just like any other interpretation, even though they are of the same tandem class, and the `"D:"` should supersede the `"C:"`.
Why do this?
Well, *in general* there is no way to know if an arbitrary tandem interpretations are of the same class.
If you create some humdrum data with new interpretations in it, `r hm` doesn't know how to parse them, so will just pile them up like this in the `Tandem` field.

Fortunately, `r hm` *does* know how to parse many standard tandem interpretations.
If you run the command 

```{r}
knownInterpretations[Type == 'Tandem'] 
```

we see that there are currently `r humdrumR:::num2word(sum(knownInterpretations$Type == 'Tandem'))` classes of tandem interpretations that `humdrumR` recognizes, each associated with a regular expression.
Thus, if we really apply `readHumdrum` to this file

```
**kern
*C:
*M4/4
*MM100
c e
c g
*D:
d f#
*-
```

it will recognize that the tokens `"C:"` and `"D:"` both match the known regular expression for the `Key` tandem interpretation, and parse them into a field called `Key`, which would look like:

```
NA
C:
C:
C:
C:
C:
D:
D:
D:
```

Likewise, we'd get both `TimeSignature` and `BPM` fields.

#### Custom Tandems

`readHumdrum` has a `tandems` argument which allows us take control of the tandem interpretation parsing process.
`tandems` is a character vector, defaulting to `known`, which parses all known interpretations into their own fields (following the `knownInterpretations` table).
We can use `tandems` two ways. 
1.) We can remove `"known"` from the vector and specify the `Name` of specific tandem interpretations we do want to parse. For instance, if we write `tandem = "Clef"`, 
the `Clef` pattern from the 
`knownInterpretations` table will be parsed, (but not any of the others).
2.) If a character string in `tandems` is not `"known"` or an exact match for a `Name` in 
`knownInterpretations`, it is instead treated as a regular expression to match a 
new tandem interpretation class.
For instance, `tandems = "[Aa]pple|[Bb]anana"` would match "\*Apple", "\*apple", "\*Banana", or "\*banana," parsing these four patterns into their own field (called `[Aa]pple|[Bb]anana]`).
We could give this field a name by writing `tandems = c(Fruit = "[Aa]pple|[Bb]anana")`.
(In fact, we can rename known interpretations as well, for instance writing `tandems = c(Meter = "TimeSignature"`.)

If `tandems = NULL`, no tandem interpretations are parsed at all---this can be used to shorten parsing time.

