---
title: "An R Primer for humdrumR"
author: "Nathaniel Condit-Schultz"
date:   "July 2022"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{An R Primer for humdrumR}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: sentence
---

```{r, include = FALSE, message=FALSE, echo = FALSE }
knitr::opts_chunk$set(
  collapse = TRUE, message = FALSE,
  comment = ">\t"
)
library(humdrumR)
```

`humdrumR` is a package for the R programming language.
You don't need to be an R master to use `humdrumR`, but there are some basic concepts from R
that you will need to learn, and ultimately, if you want to get really advanced, you'll
need to delevelop some R skills.
This document is a basic primer for R, which will teach you the basics you need to know in order to 
make the most out of `humdrumR`.


# Basic Data Structures

In R, there two fundamental data structures that are used all the time:

+ "atomic" **vectors**
+ data.tables 


## Vectors

In R, the basic units---the *atoms*, if you will---of information are called "atomic" vectors.
There are three basic atomic data types:

+ **Numbers**---`numeric` values.
  + Examples: `3`, `4.2`, `-13`, `254.30`
+ **Strings of characters**: `character` values.
  + Examples: `"note"`, `"a"`, `"do, a dear, a female dear"`
+ **Logical Trues and Falses**: `logical` values.
  + Examples: `TRUE`, `FALSE`

You might be wondering, why are we calling these basic atoms "*vectors*"?
Well, in R, the basic atomic data types are always considered a *collection* of ordered values. 
These ordered collections are called vectors.
In the simple examples above, each vector only had a single value, so it just looks like one 
value---single values like this are often called "scalars".
However, R doesn't *really* distinguish between scalars (single values) and vectors
(multiple values)---everything is always a vector.
(Still, we sometimes refer to length-1 vectors as scalars.)

To make a vector in R, use the `c` command, as so:

```{r}


c(1, 2, 3)

c("Bach", "Mozart", "Beethoven", "Brahms")

c(TRUE, FALSE)

c(32.3)

32.3

```

In this example, we've created five vectors.

+ A `numeric` vector of length 3.
+ A `character` vector of length four (composers).
+ A `logical` vector of length 2)
+ Two `numeric` vectors of length 1.
  + That's right, `c(32.3)` and `32.3` are the same thing---a vector of length 1.


---

Notice that vectors can't mix-and-match different data types; which makes sense
be cause a vector *is* a single type of thing.
But this means that commands like `c(3, "a")` will actually create
a `character` vector, where the `3` is forced to be a character (`"3"`).



### Vectorization

Having everything be a vector all the time is very useful, because it allows us to think of and 
use collections of data as single thing.
If I give you, say, ten thousand numbers, you don't have to worry about manipulating ten thousand things:
rather, you just work with *one* thing: a vector, which happens to be of length 10,000.
In R, we call this **vectorization**---generally, in R and in `humdrumR` we will constantly
be taking advantage of vectorization to make our lives super easy!

---

For an example of vectorization, watch this:

```{r}


c(1, 1, 2, 3, 5, 8, 13, 21) * 2


```

We created two `numeric` vectors:

1. The first eight numbers of the Fibonnaci sequence 
2. the single number `2`

and multiplied them together!
Notice that the entire Fibonnaci vector is multiplied by two!
We don't have to worry about multiplying each number of the vector, its done for us.

----

There are the rules for working with vectors?
There are two ideal circumstances for working with vectors.

1. They are the same length.
2. One vector is length 1, and the other isn't.


In the first case, we work with multiple vectors that are all the same length,
each value in each vector is "lined" up with values in the other vector.
If we, for example, add two such vectors together, each "lined up" pair of numbers is added:


```{r}

c(1, 2, 3) + c(5, 4, 3)


paste(c('a', 'b', 'c'), c(1, 2, 3))


```


In the second case, one of the vectors is length-1 (a "scalar").
In this case, the scalar value is paired with each value in the longer vector (as in the
Fibonnaci example above).


```{r}

c(1, 2, 3) + 5


paste(c('a', 'b', 'c'), 1)


```

----

What happens if we have vectors that are longer than one, but are not the same length.
Well, `R` will generally attempt to "recycle" the shorter vector---which means repeat it---
as necessary to match the length of the longer vector.
If the shorter vector evenly divides the longer vector, you'll generally have not problem:


```{r}


c(1, 2, 3, 4) * c(2, 3)

```

If the division is not perfect, R will still "recycle" the shorter vector, but you'll get a warning:

```{r}

c(1, 2, 3, 4) * c(2, 3, 4)

```

You see the warning message R have us?

> "longer object length is not a multiple of shorter object length"

That's R telling us that we've got an obvious mismatch in the lengths of our vectors.

---


Generally, it is best to work with vectors that are all the same length and/or scalar values (length-1 vectors),
so you can avoid worrying about how exactly R is "recycling" values.
This brings us too...


## Data tables




## Missing Data










