---
title: "An R primer for humdrumR users"
author: "Nathaniel Condit-Schultz"
date:   "July 2022"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{An R primer for humdrumR users}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: sentence
---

```{r, include = FALSE, message=FALSE, echo = FALSE }
knitr::opts_chunk$set(
  collapse = TRUE, message = FALSE,
  comment = ">\t"
)
hm <- "humdrum$_{\\mathbb{R}}$"
Hm <- "Humdrum$_{\\mathbb{R}}$"
library(humdrumR)
```

`r Hm` is a package for the R programming language.
You don't need to be an R master to use `r hm`, but there are some basic concepts from R
that you will need to learn, and ultimately, if you want to get really advanced, you'll
need to develop some R skills.
This document is a basic primer for R, which will teach you the basics you need to know in order to 
make the most out of `r hm`.


# Basic Commands

R code is made up of "expressions" like `2 + 2`, `sqrt(2)`, or `(x - mean(x))^2`.
As you can see, you can create very intuitive *arithmetic* expressions, like `5 / 2` or `3 * 3`.
However, the most common elements of R expressions are "calls" to *functions*.
A "function" in R is a pre-built bit of code that does something.
Most functions take one or more input *arguments*, and "return" some kind of output.
For example, the function `sqrt()` takes a number as an input argument, and "returns" the square root of that number.

```{r}
sqrt(2)

sqrt(9)

```

To "call" a function, we write the function's name, followed by parentheses (`()`).
Any input arguments to the function must go inside the parentheses, separated by commas if there are more than one.
Here are some examples of common functions being "called" with zero or more input arguments:

```{r}

abs(-3)

mean(1:5)

max(1, 5) # two arguments

c(1, 2, 3) # three arguments

Sys.time() # no arguments!

```


Different functions have different arguments they recognize, with specific names.
For example, the function `log()` takes two arguments, called `x` and `base`.
Other functions can take any number of arguments, with any name.
You can learn about a function, including the arguments it accepts, by typing `?functionName` at the command line; for example, `?sqrt` or `?mean`.
If you see an argument called `...`, that tells you that the function can take any number of arguments.

We can explicitly "name" the function arguments we want by putting `argname = argument` into our calls:
For example, you could say `log(10, base = 2)`.
Named arguments are very useful when we are creating data, like vectors and `data.frame`s (see below).



### Pipes

Complex expressions might involve a large number of function calls, which can get tiresome to read (or write).
For example, something like

```
log(round(sqrt(mean(x^2)), base = 2)
```

calls *four* functions!
An expression like that is a bit tricky to read, and it can be really easy to make a mistake where
you put the wrong number of parentheses.
As an alternative, R gives us the option of calling functions in a "pipe."
The way this works is we use the "pipe" command `|>`, which takes an input on the left and "pipes" it into a function call on the right.
For example, we can rewrite the previous command as:

```
x^2 |> mean() |> sqrt() |> round() |> log(base = 2)

```

Much better!
To make things even cleaner, R will understand if you spread your expressions across multiple lines, by putting a new line after each `|>`, or function argument:

```
x^2 |>
  mean() |>
  sqrt() |>
  round() |>
  log(base = 2)
  
max(sqrt(2),
    log(2),
    exp(2),
    pi / 2)

```




## Variables

When coding in R, you'll often want to "save" data or other objects so you can reuse them.
We do this by "assigning" something (often the result of a function) to a "*variable*".
This is done using the assignment operators, either `<-` or `->`.
A variable name can be any combination of upper and lowercase letters.
Let's calculate the square-root of two and save it to a variable:

```{r}
tworoot <- sqrt(2)

```

We can then reuse that value as many times as we want:

```{r}
tworoot^2

tworoot * 2

c(tworoot, tworoot)

```

You can also assign from left to right, using `->`.
This is useful in combination with pipes:

```{r}

tworoot |> exp() |> round() -> newvalue


```

----

Note your variable names can also include `_`, `.`, or numeric digits, as long as they aren't at the beginning of the name.
For example, `X1` or `my_name` are valid names---but not `2X`.





# Basic Data Structures

In R, there two fundamental data structures that are used all the time:

+ "atomic" **vectors**
+ data.frames 


## Vectors

In R, the basic units---the *atoms*, if you will---of information are called "atomic" vectors.
There are three basic atomic data types:

+ **Numbers**---`numeric` values.
  + Examples: `3`, `4.2`, `-13`, `254.30`
+ **Strings of characters**: `character` values.
  + Examples: `"note"`, `"a"`, `"do, a dear, a female dear"`
+ **Logical Trues and Falses**: `logical` values.
  + Examples: `TRUE`, `FALSE`

You might be wondering, why are we calling these basic atoms "*vectors*"?
Well, in R, the basic atomic data types are always considered a *collection* of ordered values. 
These ordered collections are called vectors.
In the simple examples above, each vector only had a single value, so it just looks like one 
value---single values like this are often called "scalars".
However, R doesn't *really* distinguish between scalars (single values) and vectors
(multiple values)---everything is always a vector.
(Still, we sometimes refer to length-1 vectors as scalars.)

To make a vector from scratch in R, use `c()`, as so:

```{r}


c(1, 2, 3)

c("Bach", "Mozart", "Beethoven", "Brahms")

c(TRUE, FALSE)

c(32.3)

32.3

```

In this example, we've created five vectors.

+ A `numeric` vector of length 3.
+ A `character` vector of length four (composers).
+ A `logical` vector of length 2.
+ Two `numeric` vectors of length 1.
  + That's right, `c(32.3)` and `32.3` are the same thing---a vector of length 1.


---

Notice that vectors can't mix-and-match different data types; which makes sense
be cause a vector *is* a single type of thing.
But this means that commands like `c(3, "a")` will actually create
a `character` vector, where the `3` is forced to be a character (`"3"`).



### Vectorization

Having everything be a vector all the time is very useful, because it allows us to think of and 
use collections of data as single thing.
If I give you, say, ten thousand numbers, you don't have to worry about manipulating ten thousand things:
rather, you just work with *one* thing: a vector, which happens to be of length 10,000.
In R, we call this **vectorization**---generally, in R and in `r hm` we will constantly
be taking advantage of vectorization to make our lives super easy!

---

For an example of vectorization, watch this:

```{r}


c(1, 1, 2, 3, 5, 8, 13, 21) * 2


```

We created two `numeric` vectors:

1. The first eight numbers of the Fibonacci sequence 
2. the single number `2`

and multiplied them together!
Notice that the entire Fibonacci vector is multiplied by two!
We don't have to worry about multiplying each number of the vector, its done for us.

----

There are two ideal circumstances for working with vectors.

1. They are the same length.
2. One vector is length 1, and the other isn't.


In the first case, we work with multiple vectors that are all the same length,
each value in each vector is "lined" up with values in the other vector.
If we, for example, add two such vectors together, each "lined up" pair of numbers is added:


```{r}

c(1, 2, 3) + c(5, 4, 3)


paste(c('a', 'b', 'c'), c(1, 2, 3))


```


In the second case, one of the vectors is length-1 (a "scalar").
In this case, the scalar value is paired with each value in the longer vector (as in the
Fibonacci example above).


```{r}

c(1, 2, 3) + 5


paste(c('a', 'b', 'c'), 1)


```

----

What happens if we have vectors that are longer than one, but are not the same length?
Well, R will generally attempt to "recycle" the shorter vector---which means repeat it---
as necessary to match the length of the longer vector.
If the shorter vector evenly divides the longer vector, you generally won't have a problem:


```{r}


c(1, 2, 3, 4) * c(2, 3)

```

If the division is not perfect, R will still "recycle" the shorter vector, but you'll get a warning:

```{r}

c(1, 2, 3, 4) * c(2, 3, 4)

```

You see the warning message R have us?

> "longer object length is not a multiple of shorter object length"

That's R telling us that we've got an obvious mismatch in the lengths of our vectors.

---


Generally, it is best to work with vectors that are all the same length and/or scalar values (length-1 vectors),
so you can avoid worrying about how exactly R is "recycling" values.
This brings us too...


## Data frames

Data frames are the heart and soul of R.
A `data.frame` is simply a collection of vectors that are *all the same length*---ideal for vectorized operations!
The vectors in a `data.frame` are arranged as columns in a two dimension table.
Let's make a data frame, by feeding some vectors to the [data.frame()] function:

```{r}

X <- c("C", "D", "E", "F", "G", "A", "B", "C")
Y <- c(0, 2, 4, 5, 7, 9, 11, 12)
Z <- c("P1", "M2", "M3", "P4", "P5", "M6", "M7", "P8")

df <- data.frame(X, Y, Z)

df

```


Notice that each of your columns/vectors can be a different type, with no problem.
Also notice, that each column has a name; we can inspect these names using the `colnames()` function.

```{r}

colnames(df)

```

Or change them:

```{r}

colnames(df) <- c('Letters', "Semitones", "Intervals")
df

```

Finally, it's also possible to assign the column name we want when creating the data frame:

```{r}
data.frame(Letters = X, Semitones = Y, Intervals = Z)

```

----

Remember, the vectors in a `data.frame` must all be the same length.
If you tried to make a `data.frame` with a vectors that don't match in length, you'll get an error "`arguments imply differing number of rows`."
The one exception is that you can call `data.frame` with some *scalar* single values, which will be automatically recycled to match
the length of the other vectors.

### With and Within You

We often want to access the columns/vectors held in a data frame.
We can do this several ways.
One approach is with the `$` operator, combined with the name of the column we want.
For example, we can get the `Letters` column from the data frame we made above using `df$Letters`.

Often, we'll want to write code that uses a bunch of different columns from the same data.frame---in fact, this is the *main* thing we do most of the time in R!
To avoid writing `df$` over and over again, we can use the `with()` function.
`with()` allows us to drop "inside" our `data.frame`, where our R commands can "see" the columns variables:

```{r}

with(df,
     paste(Intervals, Semitones, sep = ' = '))


```

## Missing Data

Sometimes we'll encounter data points which are irrelevant, meaningless, or "not applicable."


# Common Functions

+ `getwd()` --- Get R's current working directory.
+ `setwd()` --- Set R's working directory.
+ `summary()` --- Summarize the contents of an R object.

## Vector functions

+ `sort()` --- Put values of a vector into ascending order.
  + set `decreasing = TRUE` for decreasing order.
+ `rev()` --- Reverse the order of a vector.
+ `unique()` --- Returns only the unique values of a vector.
+ `x %in% y` --- Which elements of the vector `x` appear in the vector `y`?


## String functions

+ `paste()` --- Paste together multiple `character` strings.

## Arithmetic

+ `x + y` --- Addition; $x + y$.
+ `x - y` --- Subtraction; $x - y$.
+ `-x` --- Negation; $-x$.
+ `x * y` --- Multiplication; $xy$
+ `x^y` --- Exponentiation; $x^y$.
  + Use parentheses for things like `x^(1/3)`; $x^{\frac{1}{3}}$.
+ `x / y` --- Real division; $\frac{x}{y}$.
+ `x %/% y` --- [Euclidean division](https://en.wikipedia.org/wiki/Euclidean_division); $\lfloor \frac{x}{y} \rfloor$.
  + E.g., whole-number division with remainder.
+ `x %% y` --- `x` modulo `y`; $x \mod y$.
  + E.g., remainder after whole-number division.


## Other Math functions

+ `sqrt()` --- Square-root of numbers; $\sqrt{x}$.
+ `abs()` --- Absolute value of numbers; $|x|$
+ `round()` --- Round number to nearest integer; $\lfloor x \rceil$


# Making your own functions


To make your function in R, you use the `function` keyword, like so:

```
function(argument1, argument2, etc.) {
  Expressions to evaluate here, involving the arguments

}
```

For example, let's make a function that subtracts the mean from a vector of numbers.
We'll have one argument, which we'll call `numbers`.


```{r}

myfunc <- function(numbers) {
  mean <- mean(numbers)
  numbers - mean
}


```

We've created our function, and assigned it the name `myfunc`, just like any other assignment.
Let's try it out:

```{r}

myfunc(1:9)

```

Notice that the *last* expression in your function definition is the value that gets "returned" by the function.


---

If you are feeling lazy, you can also define a function using a few less keystrokes using the command `\()` instead of `function()`.
For example,

```{r}
\(x) x + 1
```
