---
title: "Working with humdrum data"
author: "Nathaniel Condit-Schultz"
date:   "July 2022"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with humdrum data}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: sentence
---

```{r, include = FALSE, message=FALSE, echo = FALSE }
knitr::opts_chunk$set(
  collapse = TRUE, message = FALSE,
  comment = ">\t"
)
library(humdrumR)
par(mar=c(1,1,1,1))
```

Once we have [imported humdrum data](ReadWrite.html "Reading and writing humdrum data") into `R`,
we'll be ready to actually work with our data!
The humdrum data you've imported is stored in a `humdrumR` data object, which gives us some special tools for working with it.

Let's load some of the prepackaged data, as we learned in the [read/write vignette](ReadWrite.html "Reading and writing humdrum data").


```{r}
chorales <- readHumdrum(humdrumRroot, 'HumdrumData/BachChorales/chor.*.krn')

chorales
```

So we've got ten files of `**kern` data.
How do we access or manipulate this data?

# With and within you

The key functions for working with `humdrumR` data objects are `with` and `within`.
These functions allow you to get "inside" your data.
Basically, once you call `with`/`within`, all the data fields within the data become available to you, to do with as you will.


## With

Let's start with `with`.
You should call `with` with a `humdrumR` data object as the first argument.
Your second argument can be any arbitrary `R` *expression* you like, and can use fields of the humdrum data.
An "expression" is just a valid bit of `R` code; for example, `2 + 2`, `c(1, 2, 3)`,  `(x - mean(x))^2`, and `nchar(letters) |> mean()` are all examples of expressions.
So "any arbitrary `R` expression" is just a fancy way of saying any `R` code that *you* want.
When we execute an `R` command, `R` "*evaluates*" the expression(s) and returns their results.

As a starting point, lets tabulate all the tokens we have in our dataset.
When you first load humdrum data, the original data tokens are always put into a field called `Token`.
We will use the *awesome* built-in `R` function, `table`, to tabulate the unique tokens in our `Token` field.
However, we can't access the `Token` field directly; if we just write the expression on the
`table(Token)` we'll get an error---`R` will not be able to "find", and will say "`Error: object 'Token' not found`." 
Instead, we use `with` to enter the data, where `R` can "see" `Token` (and all other data fields):


```{r}
with(chorales, table(Token))

```

It worked! 
However, there's a *lot* of unique tokens in this data, so its kind of a mess.
Let's take a page from the [pitch vignette](PitchAndTonality.Rmd "Pitch and tonality in humdrumR") and extract just the pitch
information using the `kern` function:

```{r}
with(chorales, 
     kern(Token) |> table())
```

Much better!
Notice that the *whole* expression `kern(Token) |> table()` is executed within the data.
Our expressions can get as complicated as we want.
Let's add a call to `sort`:

```{r}
with(chorales, 
     kern(Token) |>  table() |>  sort())
```

Ah, now we can clearly see that the top 10 most common notes in the data are `r rev(names(tail(with(chorales, sort(table(kern(Token)))), 10)))`.


---

The last step might be to make a plot.
We'll use the built in `tail` function to grab the top (last) 10, and the built-in `barplot` function.

```{r}
with(chorales, 
     kern(Token) |> 
       table() |> 
       sort() |> 
       tail(n = 10) |> 
       barplot())
```

Notice that when writing a big complex expression like 

```
 kern(Token) |> 
       table() |> 
       sort() |> 
       tail(n = 10) |> 
       barplot()
```

it's helpful to spread the expression across multiple lines. 
After the `|>` (pipe) operator is a great place for this.


## Within

In many cases, we want to apply functions to our humdrum data, but keep the humdrum/`humdrumR` structure in place.
This is what `within` is for.
Where `with` just returns what ever the output of your commands are, `within` will (attempt to) put the results of your commands
"back inside" the `humdrumR` data object, as a new field.

For example, let's say we want to translate our `kern` data into a different pitch representation, like scientific pitch.
We can do this with the `pitch` command.
If we use `with`...

```{r}
with(chorales, pitch(Token))
```

we just get all the notes!
Let's try `within` instead:

```{r}
within(chorales, pitch(Token))

```

Now we get our `humdrumR` data back, but the `Token` data has been transformed from `**kern` to `**pitch`.
Notice, that there are now two "Data fields" listed at the bottom: `Token` (the original data) and `Result1`---this is the default name for new fields produced by `within`, and yes, they will get higher numbers if you keep making new fields.

"Result1" is not very informative, so it's actually a good idea to give our new fields specific names.
We can do this by assigning our expressions in `within`:


```{r}
within(chorales, Pitch <- pitch(Token)) -> chorales

chorales

```


# With/Within Subsets

The real power of the `with` and `within` functions, comes from additional "control" arguments you can pass.
Two prominent examples allow you to automatically work with subsets of your data: the `where` and `by` arguments.
We can indicate a subset, or multiple subsets, and our expressions will 
be evaluated separately within each subset.

## With Subset

The `where` argument indicates a subset of the data which we should evaluate "within."
There `where` argument must be an expression which evaluates to a logical vector (`TRUE` of `FALSE`).
The main command is then evaluated wherever the `where` expression evaluates to `TRUE`.

Let's try it with our tabling example.
Maybe we want to table only tokens in the *first* spine:

```{r}
with(chorales, 
     kern(Token) |> table() |> sort() |> tail(n = 10),
     where = Spine == 1)
```

We only see the low notes, which is what we'd expect from the bass voice.

What about this, let's only tabulate the even-numbered bars:
For datasets which have measures indicated (with `=` tokens), the `Bar` field is integers counting these bars within each piece.
If we want even numbers, we can ask where is `Bar` modulu 2 zero?: `Bar %% 2 == 0`.

```{r}
with(chorales, 
     kern(Token) |> table() |> sort() |> tail(n = 10),
     where = Bar %% 2 == 0)
```

If we use the `where` argument with `within`, parts of the data which don't match the `within` expression
are returned as null:


```{r}

within(chorales,
       Pitch <- pitch(Token, simple = TRUE),
       where = Spine == 1)


```

Notice, our new `Pitch` field is `NULL` in all spine except spine 1.

### Or else...

In many cases when using `where` we'd like to do something with, or at least keep, the "other" part of the data.
We can do this by expressing an "or else" expression.
The idea is that we say "evaluate X where Y is true, otherwise evaluate this instead."
This is easiest to understand if we use our `wihtin` example from above.
Maybe we want to calculate the simple pitch of the first spine, but *leave the other spines unchanged*:

```{r}

within(chorales,
       Pitch <- pitch(Token, simple = TRUE),
       where = Spine == 1,
       orelse = Token)


```

This pattern, with `orelse = Token` is quite common.





## With Subgroups

The `by` keyword arguments is short for "group by."
We can indicate a field in our data to group the data by---our main call will then be applied separately to each group, and all the groups will be returned.

Let's try it with our tabling example.
Let's group the data by spine, so we tabulate each spine of data (voice) separately!

```{r}
with(chorales, 
     kern(Token) |> table() |> sort() |> tail(n = 10),
     by = Spine)
```

We get *four* different tables, one of each spine.
If we want to make multiple plots, we can just add `barplot` like before.

```{r, echo = -1}
par(mar=c(1,1,1,1))
with(chorales, 
     kern(Token) |> table() |> sort() |> tail(n = 10) |> barplot(),
     by = Spine)
```



----

Let's try something slightly different:
Since our four spines represent bass, tenor, alto, and soprano voices (in that order) I'd expect that average pitch of each spine to be significantly different.
Let's convert our pitch data to semitones (using `semits`) and add a call to the base-R function `mean`:
Watch 

```{r, echo = -1L}
par(mar=c(1,1,1,1))
with(chorales, 
     semits(Token) |> mean(),
     by = Spine)

```

Oops, that didn't work!
It's because some of our `Token` data are rests, like `4r`.
The `semits` function doesn't know how to read `4r` so it returns an `NA` value.
Don't worry, this is an easy fix, and a good lesson for data analysis in general and `R` in particular.
We always need to keep in mind what data is "missing" or Not Applicable (`NA`). 
Most `R` math functions have an option to ignore missing data by setting `na.rm = TRUE` (NA remove):

```{r}
with(chorales, 
     semits(Token) |> mean(na.rm = TRUE),
     by = Spine)

```

That's what we expected!
Of course, we might prefer a plot.

```{r}
with(chorales, 
     semits(Token) |> mean(na.rm = TRUE),
     by = Spine) |> barplot(names.arg = c('Bass', 'Tenor', 'Alto', 'Soprano'), 
                            main = 'Mean Pitch by Voice',
                            ylab = 'Semitones')



```

Notice that we did something differently here!
We don't want to make a separate plot of *each* group-by calculation---that would just be a single number for each spine.
We want to make a plot of all four numbers, so we put the `barplot` call *after* (outside) the `with` call.
Notice we can also control the labels for our barplot using the `names.arg` argument.

Maybe we want to inspect a whole histogram of the pitches in each voice.
We can use the base-R `hist` function;
Since we want a histogram of each spine, we *will* want to put `hist` "inside" the call to `with` again:


```{r, echo = -1}
par(mar=c(1,1,1,1))
with(chorales, 
     semits(Token) |> hist(xlim = c(-24, 24), main = Instrument[1]),
     by = Spine) 

```

Pro tips: by setting the `xlim` (x-limit) argument, we make sure that all four barplots cover the same ranges of pitches, so they are comparable.
We also set the main argument (figure title) using the `Instrument` field, which gives us the names of each instrument.
That's right, we can refer to any/all fields in the humdrum data!
(Not all datasets have this tandem interpretation, so it wouldn't always work.)



### Advanced Grouping

We can group our data by any arbitrary grouping in our data.
In fact the `by` keyword argument can be a complex expression, as long as its output is the same length as its input.
For example, maybe we'd like to reproduce our histograms, but lumping the male (bass/tenor) and female (alto/soprano) voices together.

```{r, echo = -1}
par(mar=c(1,1,1,1))
with(chorales, 
     semits(Token) |> hist(xlim = c(-24, 24), main = paste(unique(Instrument), sep = ' and ')),
     by = Spine < 3) 

```

The expression `Spine < 3` returns `FALSE` for spines 1 and 2 (bass and tenor) and `TRUE` for spines 3 and 4.
These two categories (`FALSE` and `TRUE`) are then used to group the data.

## Recycling results


In some cases, we'd like to perform a command `within` our data which might output a smaller vector than in the 
input, including a single value, but we still want to reconstruct the data fully.
In `R`, this is usually called "recycling"---i.e., repeating a value until it matches a certain length.
When using `with` or `within`, we can cause our results to be recycled to the full field length
by using the `fill` argument for our expression.
This is useful if you want to group by something, and fill each group with the same value:

```{r}

within(chorales,
       Semits <- semits(Token)) -> chorales

within(chorales, 
       fill = BarBassNote <- min(Semits),
       by = list(File, Bar)) -> chorales

chorales

```

What did this command do?
It looks at every bar in each file and finds the lowest note, then "fills" all the notes in that bar with that same, single, note.
We could then calculate the harmonic interval above the bar's bass note:

```{r}

within(chorales,
       Semits - BarBassNote)

with(chorales, hist(Semits - BarBassNote))

```

